Embedded Domain Specific Languages inside Functional Languages

### 2014-03-14 paper.tex
Starting conc.

In order to make deep DSLs practical, there are three folklore tricks that are used.
First, we can capture functions, by using dummy arguments. Second, we can observe loops,
buy using observable sharing. And finally, we call observe imperative statements,
by using normalization. We will overview all these tricks, before concluding
with shortcoming and remaining challenges.

@@

Expressing function calls in terms of constructors,  and
building expression trees is a useful, but by itself is a gimic. 
With careful construction, we can capture function definitions,
and other synaxtically structures,
directly from a deep embedding. It is at this
point, the idea of capturing code,
the using the captured code to execute code
on a different target becomes possible.
Consider a simple function to add one to its argument.

@@

First we need a structure for our Lava Language.
\begin{Code}

@@

instance Signal Sig where {...}

@@

The basis of a Deep EDSL is one of constructiveness.
The aim is to construct a structure, then
use that structure that was previously implicit.
Figure summarizes the state of the art with
deeply embedded reification

@@

\begin{figure}[!t]
  \centering
  \begin{minipage}{0.5\textwidth}
    \centering
    \includegraphics[width=0.8\textwidth]{images/Counter.pdf}
  \end{minipage}\begin{minipage}{0.5\textwidth}
     \centering
\footnotesize\begin{Code}[fontsize=\tiny]
entity counter is
  port(rst : in std_logic;
       clk : in std_logic;
       clk_en : in std_logic;
       restart : in std_logic;
       inc : in std_logic;
       output : out std_logic_vector(3 downto 0));
end entity counter;
architecture str of counter is
  signal sig_2_o0 : std_logic_vector(3 downto 0);
  ...
begin
  sig_2_o0 <= sig_5_o0 when (inc = '1')  else sig_6_o0;
  sig_5_o0 <= std_logic_vector(...);
  sig_6_o0 <= "0000" when (restart = '1') else sig_10_o0;
  sig_10_o0_next <= sig_2_o0;
  proc14 : process(rst,clk) is
  begin
    if rst = '1' then
      sig_10_o0 <= "0000";
    elsif rising_edge(clk) then
      if (clk_en = '1') then
        sig_10_o0 <= sig_10_o0_next;
  ....
end architecture;
\end{Code}
  \end{minipage}
  \caption{{\tt counter} Schematic and VHDL Generated by Kansas Lava}
  \label{fig:counter-pictureX}
%  \vspace{-0.1in}
\end{figure}

@@

>>>>

In order to make deep DSLs practical, there are two folklore tricks that are used almost
always used.
First, we can capture functions, by using dummy arguments. Second, we can observe loops,
buy using some form of observable sharing. 
%And finally, we call observe imperative statements, by using normalization. 
%We will overview both these tricks, before concluding with shortcomings, extensions, and remaining challenges.

@@

Expressing function calls in terms of constructors,  and
building expression trees is a useful, but by itself is a gimmick. 
However, with careful construction, we can also capture function definitions,
and other synaxtically structures,
directly from a deep embedding. It is at this
point, the idea of capturing code,
the using the captured code to execute code
on a different target becomes possible.
Consider a simple function to add one to its argument.

@@

First we need a structure for our Lava Language. We define the
functions use above, but give then a deep embedding, called \verb|Signal|.
\begin{Code}

@@

\end{Code}
Now, if we attempt to extract \verb|counter|, things go horribly wrong.
\begin{Code}
GHCi> counter (Var "restart") (Var "inc")
((( LOOPING CODE ))))
\end{Code}        
What has happened is the recursive definitions are unrolling when we try reify the function,
or more specifically, the body of counter is looping. At this point, the EDSL community was
stymied. There was efforts to use monadic structure, where the loop was expressing using
\verb|do|-notation~\cite{...}, making the loop an observable effect. There was an unsafe
extension to observe a limited form of sharing, by circumventing part of the purity of Haskell,
called observable sharing.
There was also an extension of Haskell input/output mechanism that allowed loops to
be observed indirect, called IO-based observable sharing. The net effect of all three
mechanisms is that the observed tree is rendered as a graph with named edges.

\begin{figure}[!t]
  \centering
   \begin{minipage}{0.5\textwidth}
     \centering
\footnotesize\begin{Code}[fontsize=\tiny]
entity counter is
  port(rst : in std_logic;
       clk : in std_logic;
       clk_en : in std_logic;
       restart : in std_logic;
       inc : in std_logic;
       output : out std_logic_vector(3 downto 0));
end entity counter;
architecture str of counter is
  signal sig_2_o0 : std_logic_vector(3 downto 0);
  ...
begin
  sig_2_o0 <= sig_5_o0 when (inc = '1')  else sig_6_o0;
  sig_5_o0 <= std_logic_vector(...);
  sig_6_o0 <= "0000" when (restart = '1') else sig_10_o0;
  sig_10_o0_next <= sig_2_o0;
  proc14 : process(rst,clk) is
  begin
    if rst = '1' then
      sig_10_o0 <= "0000";
    elsif rising_edge(clk) then
      if (clk_en = '1') then
        sig_10_o0 <= sig_10_o0_next;
  ....
end architecture;
\end{Code}
  \end{minipage}
  \caption{{\tt counter} Schematic and VHDL Generated by Kansas Lava}
  \label{fig:counter-pictureX}
%  \vspace{-0.1in}
\end{figure}

Haskell at this point rescues complexity. Advanced type-system mechanisms, such has higher-kind type
arguments, allow a structure to be either a tree or graph, depending on type-level instantiation. 
Omitting here the details, the reified function is a tree with sharing,
then translated into a graph with explicit sharing. The final result for our example is
\begin{Code}
GHCi> reify (counter (Var "restart") (Var "inc"))
[(0,REGISTER False 1),
 (...)
]
\end{Code}
In this output, each upper-case constructors correspond to its deep-embedding constructor.
A quick inspection shows that we have captured the circuit as given in Figure~\ref{counter-picture}.
From this netlist-style structure, it is straightforward to generate VHDL. For the example
of 4-bit number, the VHDL is given in Figure~\cite{counter-picutureX}.

These two tricks (prototypical argument, IO-based observable sharing) are the technical fundamentals
of Kansas Lava. On top of this base, and with help from Haskell type system, an entire ecosystem
for circuit generation has been developed. The DSL idiom allows programmers to use high-level 
abstraction, in Haskell, and generate efficient circuits. But not all is rosy; writing a Lava
program is not the same as writing a Haskell program because of the limitations of Deep
embeddings.

\subsection{A deep embedding is only half a program}

The basis of a Deep EDSL is one of constructiveness.
Functional programming is about constructing {\em and destructing\/} values.
Because of this, a deep embedding can not reify any pattern matching, 
or even direct usage of if-then-else, and other control flow. We side-stepped this
in Kansas Lava, for example by using a \verb|mux2| constructor, 
which encodes choice.

How much further can the idiom be pushed, if we need to be constructive?
The result is surprising. We start with our three capabilities:
\begin{itemize}
\item Basic expressions can be captured, by constructing a tree that is an analog to our syntax.
\item Functions can be captured, using the a fake unique argument.
\item Local bindings can be observed using some form of observable sharing.
\end{itemize}
With these three comes and automatic forth capability.
\begin{itemize}
\item MACRO EXPAND
\end{itemize}

Table~\cite{tab:status} gives the

The aim is to construct a structure, then
use that structure that was previously implicit.
Figure summarizes the state of the art with
deeply embedded reification

@@

@@

@@

% TODO: Something about code to generate DSL; automatic macros
